---
title: "Configure and maintain runtime hooks"
description: "Scaffold a client-owned executor, configure pre/post chains, test them offline, and regenerate safely."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Configure and maintain runtime hooks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = '#>', message = FALSE)
```

Runtime hooks customize generated wrappers before the request helper runs and
after it returns. Use them to normalize identifiers, construct service-specific
payloads, skip requests, or shape results. Your client owns the executor and the
hook functions. Installing the client does not require specmill.

This guide uses the development version of specmill with `configure_hooks()`.
Follow [installation and package setup](specmill.html) first. An older installed
version may support hook configuration without providing this setup command.
The examples below run offline when this article is built.

## Create the optional executor

For a new client, initialize the package first. For an existing package, use its
root and proceed directly to `configure_hooks()`.

```{r initialize}
schema <- system.file('catalogue/schema.json', package = 'specmill', mustWork = TRUE)
root <- tempfile('hook-client-')
specmill::initialize_client(
  root, schema, package = 'hookclient', title = 'Hook Example Client',
  author = list(given = 'Example', family = 'Maintainer', email = 'you@example.org'),
  license = 'MIT + file LICENSE', base_url = 'https://example.invalid'
)
proposal <- specmill::configure_hooks(root)
proposal$changes
proposal$missing_imports
```

Planning does not write files. The proposed executor reads YAML using `yaml`, so
add that package to the client's `DESCRIPTION` under `Imports`. Keep all existing
imports. In this disposable example we make the metadata edit programmatically:

```{r dependency}
metadata <- read.dcf(file.path(root, 'DESCRIPTION'))
metadata[1, 'Imports'] <- paste(metadata[1, 'Imports'], 'yaml', sep = ', ')
write.dcf(metadata, file.path(root, 'DESCRIPTION'))
specmill::configure_hooks(root, mode = 'apply')
```

The setup creates two files:

| File | Responsibility |
|---|---|
| `R/hook_registry.R` | Client-owned `read_hook_config()` and `run_hook()` executor |
| `inst/hooks.yml` | Ordered chains keyed by public wrapper name |

No endpoint is enabled automatically. Setup refuses to overwrite customized
files or add a second executor when it finds those function names elsewhere.
If you already have a ComptoxR-style executor, keep it and configure its callback
and YAML path instead. You do not need this scaffold.

The generated executor adds a `yaml` runtime dependency. It adds no dependency
on specmill, ComptoxR, or a separate hook framework. Commit the executor and edit
it as ordinary client code. It reads the installed YAML on each stage, without
a cache. If this becomes measurable overhead, add a cache and an explicit reload
function in the client.

## Enable pre-request and post-response hooks

In `apis/default.yml`, add these fields at service level:

```yaml
hook_callback: run_hook
hook_config: inst/hooks.yml
```

For a multi-API configuration, place them inside each participating entry under
`groups`, alongside that group's `names` and `operations`. They are not API
container or project-level fields. Each group can reference the same registry.
Hook keys are the final public R names, including any API prefix or rename.

```{r service}
service_path <- file.path(root, 'apis/default.yml')
service <- yaml::read_yaml(service_path, handlers = list(seq = function(x) x))
service$hook_callback <- 'run_hook'
service$hook_config <- 'inst/hooks.yml'
yaml::write_yaml(service, service_path)
```

Write the declarations in `inst/hooks.yml`:

```{r declarations}
writeLines(c(
  'get_item:',
  '  pre_request: [normalize_item_id]',
  '  post_response: [extract_item]'
), file.path(root, 'inst/hooks.yml'))
```

Define those functions in a separate client-owned file. This example trims the
identifier, rejects an empty identifier, and unwraps a response's `data` field.
The post-hook preserves prepared requests returned by the client's dry-run mode.

```{r functions}
writeLines(c(
  'normalize_item_id <- function(state) {',
  '  id <- trimws(state$params$item_id)',
  '  if (length(id) != 1L || is.na(id) || !nzchar(id)) stop("Supply an item ID")',
  '  state$params$item_id <- id',
  '  state',
  '}',
  '',
  'extract_item <- function(state) {',
  '  if (inherits(state$result, "httr2_request")) return(state$result)',
  '  state$result$data',
  '}'
), file.path(root, 'R/hooks_items.R'))
```

Define functions before generating wrappers. Generation validates referenced
hook names and stages. Keep these functions out of generated endpoint files.

You can instead use inline `hooks` in the service configuration:

```yaml
hook_callback: run_hook
hooks:
  get_item:
    pre_request: [normalize_item_id]
    post_response: [extract_item]
```

Choose `hooks` or `hook_config`, never both. Inline configuration tells the
**generator** which stages to emit; it does not supply the runtime executor's
registry. The scaffold reads `inst/hooks.yml`, so using that same file through
`hook_config` avoids maintaining two declarations. A custom executor may store
its policy in R instead.

## Understand the call contract

The generated flow is:

```text
public input validation
  -> run_hook(name, "pre_request", list(params = params))
  -> update public parameters from returned state$params
  -> request helper
  -> run_hook(name, "post_response", list(result = result, params = params))
  -> final return value
```

Hooks run once around the helper call. They do not run once per internal retry,
page, or batch. Transport errors propagate; `post_response` is not an error or
finally handler and does not receive failed HTTP responses automatically.

Each hook receives one argument. Pre-hooks return a list containing `params`.
Only returned parameter names matching public inputs replace those inputs.
To remove a value, retain its name with a `NULL` value, for example
`state$params['language'] <- list(NULL)`. Dropping a name leaves the original
public value unchanged. New internal request values require explicit mappings.

Chains run in declaration order, with each return value passed to the next hook.
In a post-chain, keep returning state until the final hook returns the public
result, or deliberately make later hooks accept the earlier hook's result type.
The scaffold supplies no automatic wrapping between hooks and does not inject
ComptoxR's `fn_name` or `hook_type` fields into state.

Schema-derived public inputs are validated before the pre-hook. A hook cannot
repair an input already rejected by that validation. Use explicit `inputs` and
request mappings when you need a different public contract; see
[operation settings](configuration.html#operation-settings). After transformation,
hooks own validation of their constructed values; there is no second schema
validation gate. Request-helper transport checks still apply.

## Generate and test offline

Preview, inspect the generated source, then apply:

```{r generate}
plan <- specmill::generate_client(root, config = 'specmill.yml', mode = 'plan')
plan
specmill::generate_client(root, config = 'specmill.yml', mode = 'apply')
wrapper <- readLines(file.path(root, 'R/get_item.R'))
cat(grep('run_hook|skip_request|changed|result <- api_request', wrapper, value = TRUE), sep = '\n')
```

This check loads client functions into an isolated environment and replaces the
HTTP helper with a recording function. It also points the config reader at the
source YAML. That replacement is for this source-level test only; the installed
executor uses `system.file()` to find its own package's registry.

```{r offline-test}
client <- new.env(parent = baseenv())
for (path in list.files(file.path(root, 'R'), '\\.R$', full.names = TRUE)) {
  sys.source(path, envir = client)
}
client$read_hook_config <- function() yaml::read_yaml(file.path(root, 'inst/hooks.yml'))
calls <- list()
client$api_request <- function(...) {
  calls[[length(calls) + 1L]] <<- list(...)
  list(data = list(id = 'abc', title = 'Example'))
}
stopifnot(identical(client$get_item(' abc ')$id, 'abc'))
stopifnot(identical(calls[[1]]$path_params$item_id, 'abc'))
error <- tryCatch(client$get_item('   '), error = identity)
stopifnot(inherits(error, 'client_hook_error'), length(calls) == 1L)
stopifnot(identical(error$hook_name, 'normalize_item_id'))
specmill::check_client_hooks(
  root, yaml::read_yaml(file.path(root, 'inst/hooks.yml')),
  hooks = client
)
specmill::generate_client(root, config = 'specmill.yml', mode = 'check')
```

Keep an equivalent check in your client's tests. Test chain order, representative
results, empty/invalid inputs, and any skip behavior you implement. Generated
request contract tests cover helper calls; they do not establish that your
response transformation is correct. A failed hook stops the chain and carries
its original condition in `error$parent`, plus the function, stage, and hook name.

Before distributing the package, install it and exercise the executor from the
installed namespace. This catches missing `inst/hooks.yml`, missing runtime
imports, and accidental development dependencies:

```{r installed-check, eval=FALSE}
# Run in a shell, replacing the path with your client root:
# R CMD INSTALL path/to/client
# R CMD build path/to/client
# R CMD check --no-manual hookclient_0.0.0.9000.tar.gz

library(hookclient)
stopifnot(file.exists(system.file('hooks.yml', package = 'hookclient')))
hookclient::hookclient_dry_run(TRUE)
request <- hookclient::get_item(' abc ')
stopifnot(inherits(request, 'httr2_request'))
hookclient::hookclient_dry_run(FALSE)
```

## Pass extra state or skip a request

A pre-hook can return `state$skip_request <- TRUE` and a `state$result` value.
By default the wrapper returns that result immediately, without a helper or
post-hook call. To format cached or locally computed results through the
post-chain, configure the operation:

```yaml
operations:
  GET /items/{item_id}:
    post_on_skip: true
    post_state: hook_state
```

`post_on_skip` and `post_state` are independent options. Both require a configured
pre-hook. `post_state: hook_state` passes the entire pre-state to the post-hook,
with `result` replaced by the helper result or skipped result. It is useful for
metadata such as the original identifier or a local lookup outcome.

For example, replace the pre-hook above with this version and regenerate after
adding the operation settings:

```r
normalize_item_id <- function(state) {
  state$original_id <- state$params$item_id
  state$params$item_id <- trimws(state$params$item_id)
  id <- state$params$item_id
  if (length(id) != 1L || is.na(id) || !nzchar(id)) stop('Supply an item ID')
  if (identical(state$params$item_id, 'local')) {
    state$skip_request <- TRUE
    state$result <- list(data = list(id = 'local', title = 'Local item'))
  }
  state
}
```

The skipped result deliberately has the same shape as the helper result, so
`extract_item()` handles both. The wrapper does not stop a pre-chain early when
a hook sets `skip_request`; later pre-hooks still run.

To send data constructed outside `params`, provide explicit helper mappings:

```yaml
operations:
  POST /items:
    request:
      arguments:
        method: {value: POST}
        path: {value: /items}
        path_params: {value: {}}
        query: {value: {}}
        body: {from: [hook_state, request, body]}
```

This requires a pre-hook on `create_item` that sets `state$request$body`.
`request.arguments` specifies the helper call, so include all arguments your
helper requires. For additional public options use `extra_parameters` in the
operation configuration, not ComptoxR's legacy `extra_params` inside hook YAML.
See [request bindings](configuration.html) for authentication, headers, body,
and options mappings. Hook configuration alone does not change a signature.

## Regeneration and ownership

Commit the schema, project/service YAML, runtime YAML, hook functions, executor,
generated wrappers, documentation, tests, and `.specmill/manifest.json` together.
Use `configure_hooks()` for initial setup only. It refuses to overwrite edits.
Use `generate_client()` to reconcile wrappers after configuration changes.

| Change | Required development step |
|---|---|
| Edit a hook function body | Run its tests; reinstall or reload client code |
| Reorder or replace names in an already emitted stage | Validate and regenerate/check; reinstall to ship the updated YAML |
| Add the first hook or remove the last hook for a stage | Regenerate; the wrapper must gain or lose that stage call |
| Change parameters, request mappings, skip behavior, or post-state settings | Regenerate and test the resulting helper calls and hook state |
| Rename a public wrapper | Update the hook YAML key and service names together, then regenerate |
| Update a schema | Review schema and configuration differences, regenerate, then run checks |
| Update the executor template in specmill | Review and manually adopt changes in the client-owned executor |

The executor reads the installed registry. Editing `inst/hooks.yml` in a source
checkout does not change an already installed package. Reinstall and start a
fresh R session when verifying shipped behavior. Do not edit the installed
library as your development workflow.

```r
specmill::generate_client(root, config = 'specmill.yml', mode = 'plan')
# Review the proposed changes and diagnostics.
specmill::generate_client(root, config = 'specmill.yml', mode = 'apply')
specmill::generate_client(root, config = 'specmill.yml', mode = 'check')
```

`check` makes no changes and fails for stale or protected generated output.
Normal generation retains the client-owned executor, hook functions, and runtime
YAML. If a generated file was edited by hand, move that customization into hooks
or operation configuration and follow the [ownership workflow](maintenance.html).
Do not bypass protection by deleting the manifest.

Declarations for missing wrappers in an external `hook_config` file appear in
`unused_hooks`. Review them after renaming or removing operations; they can be
intentional when several services share one registry. Referenced hook functions
must exist in client source. Generation validates wiring, not the behavior of
every possible return value.

For a ComptoxR migration, keep its existing executor, merge rules, error classes,
and state conventions. Set `hook_callback` to that executor and `hook_config` to
its registry. Specmill reads pre/post declarations from that file; legacy extra
parameters and request templates need corresponding operation configuration.
See [adopt an existing client](existing-clients.html).
