---
title: "Start a new API client"
description: "Install specmill, create a package, review generation, and verify it offline."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Start a new API client}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = '#>', message = FALSE)
```

specmill turns local API schemas and reviewed YAML policy into R wrappers,
documentation, and request contract tests. It is a **development dependency**:
people installing your generated package do not need specmill. Your package owns
HTTP requests, authentication, response parsing, and runtime hooks.

This walkthrough creates a temporary catalogue client and checks it without
contacting a service. Every evaluated example runs when this guide is built.
For an existing package, including ComptoxR, use
[Adopt an existing client](existing-clients.html) instead of initialization.

## 1. Install the development tools

This guide follows the current development version on `main`. Install it once
in R; the older `v0.1.4` release does not include all the features shown here:

```{r install, eval=FALSE}
install.packages('remotes')
remotes::install_github('seanthimons/specmill@main')
install.packages(c('httr2', 'curl', 'testthat', 'devtools'))
```

If a project supplies an installer and toolkit lock, use those instead: they
select the version reviewed for that project. In a local specmill source checkout,
`devtools::install('.')` installs the checked-out version. Start a fresh R session
after changing an already loaded version. Record the reviewed commit SHA in
your project installer or dependency lock before relying on reproducible output.
A branch name moves, and `packageVersion()` alone does not identify a development
checkout. GitHub installs also record `RemoteSha` in `packageDescription("specmill")`.

```{r version}
packageVersion('specmill')
```

You also need R package build tools appropriate to your operating system when
installing dependencies from source. The tutorial itself needs no credentials.

## 2. Choose a local schema and a new package directory

Use the bundled OpenAPI example first. It contains four operations: get an item,
list items, create an item, and refresh the catalogue.

```{r paths}
schema <- system.file('catalogue/schema.json', package = 'specmill', mustWork = TRUE)
root <- tempfile('catalogue-client-')
```

`root` is temporary so this guide can be rerun. For your own project, replace it
with an unused permanent directory such as `C:/projects/catalogueclient` or
`~/projects/catalogueclient`, and replace `schema` with your downloaded JSON, YAML, or YML
schema. Save the schema's origin and version in your project; generation reads
local files and does not download or refresh schemas.

For request and response customization, the [runtime hooks guide](hooks.html)
covers optional client-owned executor setup, examples, testing, and regeneration.

## 3. Initialize the package

Preview names, selection, and diagnostics before writing the package:

```{r configuration-preview}
proposal <- specmill::configure_client(
  root, schema, package = 'catalogueclient', name_case = 'snake_case'
)
proposal$changes
proposal$diagnostics
stopifnot(!dir.exists(root))
```

Supply your own metadata and the service's real base URL. The example URL below
is a placeholder and is never contacted in this guide.

```{r initialize}
created <- specmill::initialize_client(
  root, schema,
  package = 'catalogueclient',
  title = 'Catalogue API Client',
  author = list(given = 'Example', family = 'Maintainer', email = 'you@example.org'),
  license = 'MIT + file LICENSE',
  base_url = 'https://api.example.org',
  name_case = 'snake_case'
)
list.files(root, recursive = TRUE, all.files = TRUE)
```

Initialization writes immediately and refuses existing-file conflicts. It creates:

| File | What you maintain |
|---|---|
| `DESCRIPTION`, `LICENSE` | Package identity, dependencies, and your license |
| `R/api_request.R` | Client-owned httr2 transport for JSON/binary bodies, headers, query values, and generated authentication |
| `R/api_options.R` | Client-owned `<package>_run_verbose()` and `<package>_dry_run()` setters for package-scoped session options |
| `schema/openapi.json` | Local schema copy; YAML inputs keep `.yaml` or `.yml` |
| `.specmill/helpers/api_request.json` | Initial transport source and settings for later helper comparisons |
| `specmill.yml` | Project version and selected service files |
| `apis/<tag>.yml` (or `apis/default.yml` here) | Exact operation selection, editable names, helper, and generation policy |
| `NAMESPACE`, `.Rbuildignore` | Package exports and exclusion of development inputs |
| `.specmill/manifest.json` | Ownership hashes maintained by specmill |

The scaffold declares `httr2`, `jsonlite`, and `curl` as runtime dependencies
for requests, JSON bodies, and form uploads. Existing packages adopting the
default transport must declare these dependencies too.

Tagged schemas automatically get one service file per first operation tag, plus
grouped source files and help families. This catalogue has no tags, so it uses
`default` and keeps one source file per function. Pass `naming = 'tag_prefix'`
to propose names such as `pet_get_by_id`, or `group_by = 'none'` to keep one
service. `name_case` accepts `asis`, `snake_case`, `camel_case`, `pascal_case`,
`screaming_snake_case`, and `dot_case`. It changes proposed names only; the
editable `names` map in service YAML remains authoritative. Inspect
`attr(created, 'configuration')$diagnostics` for grouping, naming, and
unsupported-operation findings. `configure_client()` previews the same
configuration without initializing a package or overwriting existing YAML.

It does not create a Git repository, README, test runner, or pkgdown site for
the client. Those are ordinary package maintenance tasks. Do not rerun
initialization to update a client; edit policy and regenerate instead.

## 4. Inspect the schema and policy

```{r inspect}
parsed <- specmill::read_operations(file.path(root, 'schema/openapi.json'))
names(parsed$operations)
parsed$diagnostics
cat(readLines(file.path(root, 'apis/default.yml')), sep = '\n')
```

An empty diagnostics list means these operations fit the supported parser
subset. It does not prove the service works. Review diagnostics before using
real schemas; see [Supported schemas and troubleshooting](troubleshooting.html).

The service's `documentation: true` enables roxygen comments, help pages, and
exports during generation. [Configuration](configuration.html) explains how to
select methods, exclude paths, name functions, and improve their help and examples.

## 5. Preview changes before applying them

```{r plan}
plan <- specmill::generate_client(root, config = 'specmill.yml', mode = 'plan')
data.frame(
  file = vapply(plan$files, `[[`, character(1), 'file'),
  action = vapply(plan$files, `[[`, character(1), 'action')
)
stopifnot(length(plan$operations) == 4L, !length(plan$diagnostics))
```

`plan` parses and stages candidate output without changing the client. Review
the file actions, `plan$drift`, and diagnostics. `write` means a new or changed
file, `unchanged` means current output, `protected` needs your review, and
`retained` means a proposed removal is being protected. An eligible deletion
appears as `remove`. Manifest metadata can also appear in the plan.

## 6. Apply, then check reproducibility

```{r apply}
applied <- specmill::generate_client(root, config = 'specmill.yml', mode = 'apply')
checked <- specmill::generate_client(root, config = 'specmill.yml', mode = 'check')
stopifnot(all(vapply(
  checked$files, function(x) x$action %in% c('unchanged', 'retained'), logical(1)
)))
```

You now have four wrapper files in `R/`, their help pages in `man/`, and exports
in `NAMESPACE`. `check` returns normally when output is current and raises an
error when it is stale or protected. It is the default mode of
`generate_client()`. A successful check does not mean tests were run or that
every selected schema operation is supported; inspect diagnostics too.

Edit YAML policy or your client helper to change behavior. Editing a generated
wrapper makes its recorded hash differ and protects it from replacement.

## 7. Verify a wrapper without sending HTTP

For this small demonstration, source the helper and one wrapper into a private
environment, then replace the helper there with a recorder. The generated client
source has no runtime dependency on specmill.

```{r offline-call}
runtime <- new.env(parent = baseenv())
sys.source(file.path(root, 'R/api_request.R'), runtime)
sys.source(file.path(root, 'R/get_item.R'), runtime)
captured <- NULL
runtime$api_request <- function(...) {
  captured <<- list(...)
  list(id = 'item-1')
}
result <- runtime$get_item(item_id = 'item-1', language = 'en')
stopifnot(
  identical(result, list(id = 'item-1')),
  identical(captured, list(
    method = 'GET', path = '/items/{item_id}',
    path_params = list(item_id = 'item-1'), query = list(language = 'en'), body = NULL,
        request_controls = list(timeout = 30L, max_retries = 0L, retry_writes = FALSE),
        server = list(diagnostic = 'Relative server URL requires a recorded origin or explicit base URL override')
  ))
)
result
```

This checks the wrapper-to-helper contract. Follow [Write and run contract
tests](testing.html) to persist fixed expectations in the package and run them
with testthat. Transport tests are separate: they verify the final URL, headers,
serialized body, and response handling.

## 8. Install and use your client

For schemas with API-key or HTTP bearer security, initialization also proposes
an `authentication` map in `specmill.yml`. Generation creates `R/api_auth.R` with
`set_api_token()` and `api_token()`. Configure the token at runtime; keep it out
of YAML. Public endpoints need no token. OAuth login and refresh are deferred.
See the [configuration guide](configuration.html) for setup and persistence.

After configuring the actual service and any required credentials, run:

```{r client-install, eval=FALSE}
devtools::install(root)
help('get_item', package = 'catalogueclient')
catalogueclient::catalogueclient_dry_run(TRUE)
request <- catalogueclient::get_item(item_id = 'a-real-id', language = 'en')
catalogueclient::catalogueclient_dry_run(FALSE)
# This call sends HTTP:
catalogueclient::get_item(item_id = 'a-real-id', language = 'en')
```

Dry runs still validate inputs and require credentials for authenticated
operations. The returned request may contain credentials; do not publish it.
Use `catalogueclient_run_verbose(TRUE)` to report the method and response status,
and `catalogueclient_run_verbose(FALSE)` to turn messages off.

The last call sends a real request. The default helper performs one request,
including when an argument is named `page`. It returns JSON objects and arrays
as R lists, JSON scalars as their corresponding R scalars, text/SVG as strings,
binary as raw bytes, and empty bodies or JSON `null` as `NULL`. It raises
errors for HTTP failures or malformed nonempty JSON. Those errors report the
status and response media type without including the request URL, response body,
or credential values. A missing or unrecognized `Content-Type` returns raw bytes.
A JSON `Content-Type` with a non-JSON body is a decode error, while a text type
always returns text even if its body happens to contain JSON. The parser records
declared response media and schema metadata, but runtime decoding follows the
actual `Content-Type`. The generator does not pass declared response schemas to
the helper or validate response bodies against them. Such validation would need
an explicit generated helper argument and remains unsupported. Credentials,
pagination, batching, and service-specific validation belong in the client.
Optional `paginated()` companions support explicitly configured page/offset
and cursor retrieval, with `paginated_links()` for next links; see [pagination configuration](configuration.html#pagination).
New helpers expose timeout, URL override, and bounded safe retries through one
package-scoped R option; see the [request controls guide](https://github.com/seanthimons/specmill#generated-client-request-controls).
Existing client-owned helpers require an explicit update to adopt these controls.

## 9. Save the project and maintain it

Commit the schema, YAML policy, client helper, generated source/help/tests,
`.specmill/manifest.json`, and `.specmill/helpers/` together. Keep the manifest tracked: it establishes which
output can be updated safely. Keep credentials outside tracked files. Add a
package test runner as shown in the testing guide.

When a schema or policy changes, repeat **plan -> review -> apply -> check ->
tests**. See [Routine maintenance and CI](maintenance.html) for thin commands,
schema comparisons, coverage reports, and building documentation for your client.
